Analyzing the Privacy Paradox Using a Nationally Representative Three-Wave Panel Study

### 2020-10-29 manuscript.RmD
- implemented last minor edits (mainly language)

---
title             : "Analyzing the Privacy Paradox Using a Nationally Representative Three-Wave Panel Study" 
shorttitle        : "Analyzing the Privacy Paradox"
abstract: |
  The privacy paradox states that people’s concerns about online privacy are unrelated to their online sharing of personal information. On the basis of a representative sample of the German population, which includes `r knitr::load_cache(label = "load-data", object = "n_complete")` respondents interviewed at three waves separated by 6 months, we investigate the privacy paradox from a longitudinal perspective. Using a cross-lagged panel model with random intercepts, we differentiate between-person relations from within-person effects. Results revealed that people who were more concerned about their online privacy than others also shared slightly less personal information and had substantially more negative attitudes toward information sharing (between-person level). People who were more concerned than usual also shared slightly less information than usual (within-person level). We found no long-term effects of privacy concerns on information sharing or attitudes 6 months later. The results provide further evidence against the privacy paradox, but more research is needed to better understand potential causal relations.
keywords          : "privacy paradox, privacy concerns, information sharing, longitudinal analysis, structural equation modeling"
wordcount         : "`r wordcountaddin::word_count('manuscript.Rmd')`"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
class             : "man"
mask              : yes
# to compile as pdf
# (a) move file "_site.yml" to folder "tmp"
# (b) uncomment the following lines until incl. "\raggedbottom"
# (c) comment lines after "output: html_document ..."
# output            : papaja::apa6_pdf
# header-includes:
#   - \setlength{\parskip}{0em}
#   - \raggedbottom
output:
  html_document:
    toc: TRUE
    toc_float:
      collapsed: FALSE
    toc_depth: 2
bibliography      : 
  - [bibliography/references.bib]
  - [bibliography/r-references.bib]
csl               : "sage-harvard.csl"
---

@@

# Introduction

@@

With this study we aim to answer four major questions. 
First, on a between-person level, how are concerns about online privacy related to the online sharing of personal information? 
Second, on a within-person level, does information sharing decrease when concerns increase?
Third, what are the potential long-term effects?
Are changes in concerns related to changes in information sharing 6 months later and/or vice versa? 
Fourth, what is the role of privacy attitudes, do they mediate the relation between privacy concerns and information sharing?
To best answer and contextualize these questions, we first provide an in-depth theoretical analysis of the privacy paradox, after which we present the empirical results of a longitudinal panel study, which is representative of the German population.

@@

In general, we believe that it should be possible to find long-term effects. 
It has been argued that privacy concerns affect privacy behavior in the long run [e.g., @heirmanPredictingAdolescentsDisclosure2013]. 
The underlying theoretical mechanism could be that the emotional part of privacy concerns causes (a) motivated information selection and (b) motivated information processing, which is likely to change actual behavior [@nabiCognitivefunctionalModelEffects1999]. 
Specifically, when privacy concerns increase (e.g., because of experienced or witnessed privacy infringements), people might begin reading more media articles on privacy issues and might also consume these articles more carefully, which could prompt information sharing practices that are more cautious.
Also empirically, a study with 290 participants found small negative longitudinal (between-person) relations between privacy concerns and self-disclosure [@koohikamaliInvestigationDynamicModel2019].

@@

The results of the between-person analyses showed that people who were more concerned about their privacy than others were slightly less likely to share personal information.
In addition, people who were more concerned about their privacy than others also held substantially more negative attitudes toward disclosing personal information online.
Notably, we found a very strong between-person correlation between attitudes toward information sharing and actual information sharing, which shows that typical online disclosure can be predicted precisely by a person's attitude.
Taken together, the cross-sectional results are in line with the extant literature:
The between-person correlation of privacy concerns and information sharing found in this study (i.e., $\beta$ = -.09) fall within the 95% confidence interval of the effect reported by @baruhOnlinePrivacyConcerns2017 (i.e., _r_ = -.13, 95% CI [-.07, -.18]). 
Note that the between-person correlations reported here represent averaged measurements across three waves, which makes the findings more robust than typical one-shot measures.

@@

The within-person results showed that when a person's privacy concerns increased, the same person also shared slightly less information online than usual.
Moreover, people who developed more positive attitudes toward the online sharing of personal information than usual, also shared substantially more personal information online.
Together, changes in concerns and attitudes are therefore related to changes in behavior, which speaks against the privacy paradox also on the within-person level.

@@

The data were collected between May 2014 and May 2015---hence, after the Snowden revelations in 2013, but before the Equifax data breach in 2017, Cambridge Analytica (2018), or the implementation of the General Data Protection Regulation (2018). 
Such sweeping events, however, could affect privacy concerns, online behavior, or their mutual relation, which would limit the generalizability of our results. 
Although this is an important caveat, we have reason to believe that our findings are largely robust. 
First, additional analyses showed that the within-person relationships were stable across waves (a period of 1 year). 
Second, another set of additional analyses showed that most effects remained stable until winter 2017.
Third, records of online search terms revealed that although interest in privacy-related topics and privacy-enhancing technologies increased after the Snowden revelations, it returned to prior levels after only two weeks [@preibuschPrivacyBehaviorsSnowden2015]. 
It thus seems that levels of privacy concerns and information sharing, as well as their mutual relationship, are largely robust.

>>>>

---
title             : "Analyzing the Privacy Paradox Using a Nationally Representative Three-Wave Panel Study" 
shorttitle        : "Analyzing the Privacy Paradox"
abstract: |
  The privacy paradox states that people’s concerns about online privacy are unrelated to their online sharing of personal information. On the basis of a representative sample of the German population, which includes `r knitr::load_cache(label = "load-data", object = "n_complete")` respondents interviewed at three waves separated by 6 months, we investigate the privacy paradox from a longitudinal perspective. Using a cross-lagged panel model with random intercepts, we differentiate between-person relations from within-person effects. Results revealed that people who were more concerned about their online privacy than others also shared slightly less personal information and had substantially more negative attitudes toward information sharing (between-person level). People who were more concerned than usual also shared slightly less information than usual (within-person level). We found no long-term effects of privacy concerns on information sharing or attitudes 6 months later. The results provide further evidence against the privacy paradox, but more research is needed to better understand potential causal relations.
keywords          : "privacy paradox, privacy concerns, information sharing, longitudinal analysis, structural equation modeling"
wordcount         : "`r wordcountaddin::word_count('manuscript.Rmd')`"
figsintext        : yes
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
class             : "man"
mask              : yes
# to compile as pdf
# (a) move file "_site.yml" to folder "tmp"
# (b) uncomment the following lines until incl. "\raggedbottom"
# (c) comment lines after "output: html_document ..."
output            : papaja::apa6_pdf
header-includes:
  - \setlength{\parskip}{0em}
  - \raggedbottom
# output:
#   html_document:
#     toc: TRUE
#     toc_float:
#       collapsed: FALSE
#     toc_depth: 2
bibliography      : 
  - [bibliography/references.bib]
  - [bibliography/r-references.bib]
csl               : "sage-harvard.csl"
---

@@

<!-- # Introduction -->

@@

With this study we aim to answer four major questions. 
First, on a between-person level, how are concerns about online privacy related to the online sharing of personal information? 
Second, on a within-person level, is information sharing lower than usual when concerns are higher than usual?
Third, what are the potential long-term effects?
Are changes in concerns related to changes in information sharing 6 months later and/or vice versa? 
Fourth, what is the role of privacy attitudes, do they mediate the relation between privacy concerns and information sharing?
To best answer and contextualize these questions, we first provide an in-depth theoretical analysis of the privacy paradox, after which we present the empirical results of a longitudinal panel study, which is representative of the German population.

@@

In general, we believe that it should be possible to find long-term effects. 
It has been argued that privacy concerns affect privacy behavior in the long run [e.g., @heirmanPredictingAdolescentsDisclosure2013]. 
The underlying theoretical mechanism could be that the emotional part of privacy concerns causes (a) motivated information selection and (b) motivated information processing, which is likely to change actual behavior [@nabiCognitivefunctionalModelEffects1999]. 
Specifically, when privacy concerns are higher than usual (e.g., because of experienced or witnessed privacy infringements), people might begin reading more media articles on privacy issues and might also consume these articles more carefully, which could prompt information sharing practices that are more cautious.
Also empirically, a study with 290 participants found small negative longitudinal (between-person) relations between privacy concerns and self-disclosure [@koohikamaliInvestigationDynamicModel2019].

@@

The results of the between-person analyses showed that people who were more concerned about their privacy than others were slightly less likely to share personal information.
In addition, people who were more concerned about their privacy than others also held substantially more negative attitudes toward disclosing personal information online.
Notably, we found a very strong between-person correlation between attitudes toward information sharing and actual information sharing, which shows that typical online disclosure can be predicted precisely by a person's attitude.
Taken together, the cross-sectional results are in line with the extant literature:
The between-person correlation of privacy concerns and information sharing found in this study (i.e., $\beta$ = -.09) fall within the 95% confidence interval of the effect reported by @baruhOnlinePrivacyConcerns2017 (i.e., _r_ = -.13, 95% CI [-.07, -.18]). 
Notably, the between-person correlations reported here represent averaged measurements across three waves, which makes the findings more robust than typical one-shot measures.

@@

The within-person results showed that when a person's privacy concerns are higher than usual, the same person also shared slightly less information online than usual.
Moreover, people who developed more positive attitudes toward the online sharing of personal information than usual, also shared substantially more personal information online.
Together, changes in concerns and attitudes are therefore related to changes in behavior, which speaks against the privacy paradox also on the within-person level.

@@

The data were collected between May 2014 and May 2015---hence, after the Snowden revelations in 2013, but before the Equifax data breach in 2017, the Cambridge Analytica data breach in 2018, or the implementation of the General Data Protection Regulation in 2018. 
Such sweeping events, however, could affect privacy concerns, online behavior, or their mutual relation, which would limit the generalizability of our results. 
Although this is an important caveat, we have reason to believe that our findings are largely robust. 
First, additional analyses showed that the within-person relationships were stable across waves (a period of 1 year). 
Second, another set of additional analyses showed that most effects remained stable until winter 2017.
Third, records of online search terms revealed that although interest in privacy-related topics and privacy-enhancing technologies increased after the Snowden revelations, it returned to prior levels after only two weeks [@preibuschPrivacyBehaviorsSnowden2015]. 
It thus seems that levels of privacy concerns and information sharing, as well as their mutual relationship, are largely robust.
